Compositional Verification of Concurrent C Programs with Search Structure Templates

### 2023-11-24 concurrency/CPP 2024/paper.tex
WIP

\lstdefinestyle{myStyle}{
	backgroundcolor=\color{white},   
	commentstyle=\color{codegreen},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	keepspaces=true,                 
	numbers=left,       
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=1,
}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{Concurrent separation logic, fine-grained locking, Iris, logical atomicity, theorem proving, Verified Software Toolchain}

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\begin{lstlisting}[language=caml, style=myStyle]
			let insert r k =
			  lockNode r;
			  let n = traverse r k in
			  let res = insertOp n k in
			  unlockNode n; 
			  res\end{lstlisting}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The lock-coupling search structure template}
	\label{template-ex}
\end{figure}

\begin{figure*}

\end{figure*}

Figure~\ref{template-ex} shows an example search structure template. The core of the template is the \lstinline{traverse} function, which uses a specific concurrency control mechanism to travel through a data structure in search of the requested key. The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine. Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node. Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics. The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure. Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *left, *right;
	} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *next;
	} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

The first template we consider is lock coupling (also called hand-over-hand locking), in which threads use the locks on each node to prevent interference from other threads during traversal. Each thread always holds at least one lock, and acquires the lock on the next node before releasing its current lock, ensuring that other threads cannot invalidate the ongoing search.
Figure \ref{traverse_lock} shows the lock-coupling \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and our corresponding C implementation (Figure \ref{traverse_lock_b}). The C implementation uses a struct
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, xleftmargin=-2.07em]
	typedef struct pn {
		struct node_t *p; struct node_t *n;
	} pn;
\end{lstlisting}
to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent. The template relies on one function provided by the underlying data structure, namely \lstinline{findNext}, which is used to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$. In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a Boolean and, if a next node is found, modifies \lstinline{pn->n}.

@@

\begin{figure*}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
			\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
			\label{traverse_lock_a}	
		\end{subfigure}
		\\ \\ 
		\renewcommand{\thesubfigure}{c}% New fixed/manual numbering
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=C, style=myStyle]{lock_insert.c} 
			\caption{The \lstinline{insert} method of the lock-coupling template algorithm written in C}
			\label{insert_lock}	
		\end{subfigure}
	\end{subfigure}\qquad
	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms 
		%\wm{I like the layout here, but it would make more sense if \lstinline{traverse} was b and \lstinline{insert} was c}
	}
	\label{traverse_lock}
\end{figure*}

@@

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ \ast \ \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}

\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

\begin{figure*}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
			\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
			\label{traverse_giveup_a}	
		\end{subfigure}
		\\ \\ 
		\renewcommand{\thesubfigure}{c}% New fixed/manual numbering
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=C, style=myStyle]{giveup_insert.c} 
			\caption{The \lstinline{insert} method of the give-up template algorithm written in C}
			\label{insert_giveup}	
		\end{subfigure}
	\end{subfigure}\qquad
	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
	\begin{subfigure}[t]{0.48\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} and \lstinline{insert} methods of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure*}

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min, max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	} \vspace{-0.85em} 
	\\ \texttt{inRange(pn, k)} 
	\vspace{-0.85em}  \\
	{\color{blue}
		\left\{\mathit{res.} 
		\begin{array}{l}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  \ 
			\\\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

\subsubsection{Verifying the Give-Up Template}
\label{traverse_proof_giveup}

The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{blue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ \ast \ 
				\nodeboxrep(n)   
			\end{array}  
			\Big \vert  \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{blue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast\ \texttt{k} \in \mathsf{range} \ \ast \ 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}  
			\Bigg \vert \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

@@

\subsection{Instantiating \lstinline{findNext} and \lstinline{insertOp}}
We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}). As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.

\begin{figure*}[!ht]
	\centering
	\begin{mathpar}
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)
			\end{array}
			\right\}
		} \vspace{-0.85em} 
		\\ 
		\texttt{findNext(pn, k)} 
		\vspace{-0.85em} 
		\\
		{\color{blue}
			\left\{\mathit{res}. \ \exists \  n'.
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure*}

>>>>

\lstdefinestyle{myStyle}{
	backgroundcolor=\color{white},   
	commentstyle=\color{codegreen},
	keywordstyle=\bfseries\color{green!40!black},
	commentstyle=\itshape\color{purple!40!black},
	stringstyle=\color{codepurple},
	basicstyle=\ttfamily\footnotesize,
	breakatwhitespace=false,         
	breaklines=true,                 
	keepspaces=true,                 
	numbers=left,       
	numbersep=5pt,                  
	showspaces=false,                
	showstringspaces=false,
	showtabs=false,                  
	tabsize=2,
}

@@

%%
%% Keywords. The author(s) should pick words that accurately describe
%% the work being presented. Separate the keywords with commas.
\keywords{concurrent separation logic, fine-grained locking, Iris, logical atomicity, interactive theorem proving, Verified Software Toolchain}

@@

\begin{figure}[h]
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
		%		\label{traverse_lock_a}	
	\end{subfigure}\qquad
	\begin{subfigure}[t]{0.45\textwidth}
		\lstinputlisting[language=caml, style=myStyle]{insert.ml} 
	\end{subfigure}
		%		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		%		\label{traverse_lock_b}
	\caption{The lock-coupling search structure template}
	\label{template-ex}
\end{figure}
Figure~\ref{template-ex} shows an example search structure template. The core of the template is the \lstinline{traverse} function, which uses a specific concurrency control mechanism to travel through a data structure in search of the requested key. The mechanism in this example is \emph{lock coupling}, where we acquire the lock on the next node before releasing the lock on the current node. The node to travel to is selected by a black-box function \lstinline{findNext} provided by the data structure; all the template needs to know is that it has some way of choosing a next node to examine. Once the appropriate node for the key has been found, the template returns it to a top-level function such as \lstinline{insert} that calls out to the data structure to perform the actual insertion on the node. Thus, the \lstinline{traverse} and \lstinline{insert} functions can be verified without knowing anything about the target data structure other than its synchronization mechanism, as long as the data structure implements \lstinline{findNext} and \lstinline{insertOp} operations with the required semantics. The concurrency functions in the template and the sequential \lstinline{findNext} and \lstinline{insertOp} functions provided by the data structure combine into a fully operational concurrent data structure. Krishna et al.~specify templates in terms of \emph{flow interfaces}~\cite{krishna2017flow}, a framework for reasoning about graph-structured data structures (i.e., those in which a node may be reachable along multiple paths), but many common search structures do not require flow-style reasoning; in this paper, we present a version of templates that is independent of flow interfaces. %is this a contribution in itself?
%what more do we want to say here?

@@

The core of a search structure template is a \lstinline{traverse} function that navigates a data structure using the chosen synchronization mechanism. A template also includes wrapper functions that use \lstinline{traverse} to perform data structure operations (insert, lookup, etc.), and may include helper functions as well. The defining feature of all these functions is that they abstract away from the data structure being traversed: they operate on a generic \lstinline{node} type and take data-structure-specific functions like \lstinline{findNext} (which chooses the next node to search) as parameters. We consider lock-based synchronization in this paper, and so our generic node type can be defined in C as:
\begin{lstlisting}[style=myStyle,language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
	typedef struct node_t {node *t; lock_t *lock;} node_t;
\end{lstlisting}
There are two things worth noting about this data structure. First, each node has its own lock: all of our templates will use \emph{fine-grained locking}, holding locks on as few nodes as possible to allow other parts of the data structure to be modified concurrently. Second, the \lstinline{node} struct itself is defined by the target data structure: for instance, a binary search tree might define it as 
\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *left, *right;
	} node;
\end{lstlisting}
while a linked list might use
\begin{lstlisting}[style=myStyle, numbers=none, xleftmargin=0.5em]
	typedef struct node {
		int key; 
		void *value; 
		struct node_t *next;
	} node;
\end{lstlisting}
This is the first place where we decompose our program into a generic concurrent component and a sequential data-structure component. Next, we use \lstinline{node_t} to define template functions for specific concurrency patterns, and then prove (assuming appropriate specifications for data-structure-specific functions) that they correctly implement the data structure operations.

@@

The first template we consider is lock coupling (also called hand-over-hand locking), in which threads use the locks on each node to prevent interference from other threads during traversal. Each thread always holds at least one lock, and acquires the lock on the next node before releasing its current lock, ensuring that other threads cannot invalidate the ongoing search.
Figure \ref{traverse_lock} shows the lock-coupling \lstinline{traverse} function as presented by Krishna et al.~(Figure \ref{traverse_lock_a}) and our corresponding C implementation (Figure \ref{traverse_lock_b}). The C implementation uses a struct
\begin{lstlisting}[style=myStyle, language = C, backgroundcolor=\color{white}, basicstyle=\ttfamily\footnotesize, numbers=none, xleftmargin=0.5em]
	typedef struct pn {
		struct node_t *p; 
		struct node_t *n;
	} pn;
\end{lstlisting}
to mimic the pair of nodes \lstinline{(p, n)} returned by the functional implementation, where \lstinline{n} is the current node and \lstinline{p} is its parent. The template relies on one function provided by the underlying data structure, namely \lstinline{findNext}, which is used to determine the next node $\texttt{n'}$ to be visited based on the current node $\texttt{n}$ and the key $\texttt{k}$. In the functional version, \lstinline{findNext} returns an \lstinline{option node}; in C, it instead returns a Boolean and, if a next node is found, modifies \lstinline{pn->n}.

@@

\begin{figure*}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{lock_traverse.ml} 
			\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in an ML-like language \cite{krishna2019compositional}} 
			\label{traverse_lock_a}	
		\end{subfigure}
		\\ \\ 
		\renewcommand{\thesubfigure}{c}% New fixed/manual numbering
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[style=myStyle]{lock_insert.c} 
			\caption{The \lstinline{insert} method of the lock-coupling template algorithm written in C}
			\label{insert_lock}	
		\end{subfigure}
	\end{subfigure}\qquad
	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
	\begin{subfigure}[t]{0.44\textwidth}
		\lstinputlisting[language=C, style=myStyle]{lock_traverse.c} 
		\caption{The \lstinline{traverse} method of the lock-coupling template algorithm written in C}
		\label{traverse_lock_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} method of the lock-coupling template algorithms 
		%\wm{I like the layout here, but it would make more sense if \lstinline{traverse} was b and \lstinline{insert} was c}
	}
	\label{traverse_lock}
\end{figure*}

@@

The resources $\mathsf{R}$ contained in a node depend on the specific data structure, but always include a piece of \emph{ghost state} describing the current state of the node (its key, value, and key range) shared between the invariant $\mathsf{R}$ and the abstract state $\treerep$, ensuring that the lock and the abstract state agree on the contents of the node. Formally, the lock invariant is defined by 
\begin{align*}&\mathsf{node\_contents}(n, c, \mathsf{range}) \triangleq \\ 
	&\qquad\mathsf{ghost\_node}(n, c,\mathsf{range}) \ast \mathsf{node\_data}(n, c) \\
	&\mathsf{R}(n) \triangleq \exists\ c, \mathsf{range}.\ \mathsf{node\_contents}(n, c, \mathsf{range})\end{align*}
where the definition of $\mathsf{node\_data}$ is supplied by the target data structure. The contents $c$ of a node can be either a key-value pair $(k, v)$, or the empty contents $\cdot$ (used for nodes that have been allocated but not yet assigned keys). Then $\treerep\ (m)$ is defined as a collection of $\mathsf{ghost\_node}$s that form a tree containing all the key-value pairs in $m$.

@@

The key to the correctness of the \texttt{traverse} function is the loop invariant for the top-level loop, which expresses that in each iteration, \lstinline{traverse} holds the lock on a node that has \lstinline{k} in its range:
\begin{align*}\mathsf{traverse\_inv}(\texttt{pn}) \triangleq \\ \exists \ p, n, c, \mathsf{range}.\ &\texttt{pn} \mapsto (p, n) \ast \texttt{k} \in \mathsf{range}\ \ast \ \\  \nodeboxrep(n)  \ \ast \  &\mathsf{node\_contents}(n, c, \mathsf{range})
\end{align*}
\begin{figure*}[!ht]
	$\color{blue}
	\fforall \  m. \left\langle \texttt{pn} \mapsto (p, n) \ \ast \ 
	\nodeboxrep(n)  \ \ast \ \mathsf{R}(n) \ \big| \ \treerep\ (m) \
	\right\rangle$
	\lstinputlisting[language=C, style=myStyle, mathescape=true]{proof_lock_traverse.c}
	$\color{blue}
	\left\langle \mathit{res}. \ \exists \  n', v.
	\begin{array}{l} \texttt{pn} \mapsto (n', n') \ \ast \nodeboxrep(n') \ \ast\ \texttt{k} \in \mathsf{range}\ \ast\ 
		\\ 
		\mathsf{if} \ \mathit{res} \ \mathsf{then} \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
		\\ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
	\end{array}
	\ \Bigg| \treerep\ (m) \
	\right\rangle$
	\caption{Proof outline of the lock-coupling \texttt{traverse} function}
	\label{proof_lock_traverse}
\end{figure*}

@@

\begin{figure*}[h]
	\begin{subfigure}[t]{0.48\textwidth}
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[language=caml, style=myStyle]{giveup_traverse.ml} 
			\caption{The \lstinline{traverse} method of the give-up template algorithm written in an ML-like language}
			\label{traverse_giveup_a}	
		\end{subfigure}
		\\ \\ 
		\renewcommand{\thesubfigure}{c}% New fixed/manual numbering
		\begin{subfigure}[t]{\textwidth}
			\lstinputlisting[style=myStyle]{giveup_insert.c} 
			\caption{The \lstinline{insert} method of the give-up template algorithm written in C}
			\label{insert_giveup}	
		\end{subfigure}
	\end{subfigure}\qquad
	\renewcommand{\thesubfigure}{b}% New fixed/manual numbering
	\begin{subfigure}[t]{0.44\textwidth}
		\lstinputlisting[language=C, style=myStyle]{giveup_traverse.c} 
		\caption{The \lstinline{traverse} method of the give-up template algorithm written in C}
		\label{traverse_giveup_b}
	\end{subfigure}
	\caption{The \lstinline{traverse} and \lstinline{insert} methods of the give-up template algorithms}
	\label{traverse_giveup}
\end{figure*}

@@

The \lstinline{inRange} function raises an interesting question about the template approach: is \lstinline{inRange} part of the give-up template, or the underlying data structure? Some data structures may already track the range of keys expected in the current node, and so might define \lstinline{inRange} even in sequential implementations. However, in most sequential settings ranges can be computed from the data structure (e.g., in a binary search tree, the left subtree of a node with key \lstinline{k} holds keys less than \lstinline{k}), and there is no reason to explicitly store a node's range in the node itself. The give-up template of Krishna et al.~implicitly assumes that the underlying data structure supports \lstinline{inRange}; we prefer to consider \lstinline{inRange} part of the template, so that the template can be applied to data structures without modifying them. Accordingly, in this template we add two new fields, \lstinline{min} and \lstinline{max}, to the type \lstinline{node_t}:
\begin{lstlisting}[style=myStyle,  numbers=none, xleftmargin=0.5em]
	typedef struct node_t {
		node *t; 
		lock_t *lock; 
		int min, max;
	} node_t;
\end{lstlisting}
These fields store lower and upper bounds on the keys reachable from the current node. %In the context of binary search trees, each node is labeled with a range that is associated with a lower and upper bound on the keys. This begins with $(\texttt{min}, \texttt{max}) = (-\infty, +\infty)$ at the root node and is propagated to the empty leaf nodes. If the root node (which also serves as the parent of a node) has a key of $42$, the left leaf node's range would be $(\texttt{min}, \texttt{max}) = (-\infty, 42)$, and the right leaf node's range would be $(\texttt{min}, \texttt{max}) = (42, +\infty)$.
We can then define \texttt{inRange} as a helper function in the template, rather than requiring data structures to provide it. %\wm{This is the first but not the last implicit assumption in the templates; can we say something more general?}
Our specification of \lstinline{inRange} is:
\begin{mathpar}
	% inRange for give-up
	{\color{blue}
		\left\{ 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2
		\end{array}
		\right\}
	} \vspace{-0.85em} 
	\\ \texttt{inRange(pn, k)} 
	\vspace{-0.85em}  \\
	{\color{blue}
		\left\{\mathit{res.} 
		\begin{array}{c}
			\texttt{pn->n->min}\mapsto n_1 \ast \texttt{pn->n->max}\mapsto n_2 \ \ast  
			\\ \mathsf{if}\ \mathit{res}\ \mathsf{then}\ (n_1 < \texttt{k} < n_2)\ \mathsf{else}\ (\texttt{k} \leq n_1 \lor \texttt{k} \geq n_2)
		\end{array}
		\right\}
	}
\end{mathpar}
It simply computes whether the input key \lstinline{k} is in the range $(n_1, n_2)$ associated with the node \texttt{n}, and returns 0 or 1 accordingly.

\subsubsection{Verifying the Give-Up Template}
\label{traverse_proof_giveup}
The give-up template's \texttt{traverse} specification is almost the same as in the lock-coupling template, except that the caller does not need to hold any locks before calling it, so the invariant $\mathsf{R}$ does not appear in the precondition:
\begin{mathpar}
	{ \color{blue}
		\fforall \  m. \ \scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\texttt{pn} \mapsto (p, n) \ast  
				\nodeboxrep(n)   
			\end{array}  
			\Big \vert  \treerep\ (m)
		}  
		{\rangle}
	}
	\vspace{-6pt}\\
	\vspace{-6pt}\texttt{traverse(pn, k)} \\
	{ \color{blue}
		\scaleleftright[1.4ex]{\langle}
		{\begin{array}{l}
				\mathit{res}. \exists \  n', v, \mathsf{range}. \\ 
				\ \ \ \ \ \  \texttt{pn} \mapsto \left(n', n'\right) \ast  \nodeboxrep\left(n'\right) \ast \texttt{k} \in \mathsf{range} \ \ast 
				\\ 
				\ \ \ \ \ \ \mathsf{if} \ \mathit{res}  \ \mathsf{then}  \ \mathsf{node\_contents}(n', \cdot, \mathsf{range}) \ 
				\\ 
				\ \ \ \ \ \ \ \ \ \ \ \ \ \ \ \mathsf{else} \ \mathsf{node\_contents}(n', (\texttt{k}, v), \mathsf{range})
			\end{array}  
			\Bigg \vert \treerep\ (m)
		}  
		{\rangle}
	}
\end{mathpar}

@@

\subsection{Instantiating \lstinline{findNext} and \lstinline{insertOp}}
We instantiate the templates with the binary search tree by defining and verifying the helper functions \texttt{findNext} (used in \lstinline{traverse}) and \texttt{insertOp} (used in \texttt{insert}). As expected, their implementations are strictly sequential, their specifications are ordinary non-atomic Hoare triples (see Figures \ref{fig:findNext_lock} and \ref{fig:insertOp}), and their proofs do not require any concurrency reasoning.
\begin{figure*}[!ht]
	\centering
	\begin{mathpar}
		{\color{blue}
			\left\{ 
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)
			\end{array}
			\right\}
		} \vspace{-0.85em} 
		\\ 
		\texttt{findNext(pn, k)} 
		\vspace{-0.85em} 
		\\
		{\color{blue}
			\left\{\mathit{res}. \ \exists \  n'.
			\begin{array}{c}
				\texttt{pn->n->t}\mapsto (k', v', l, r)  \ \ast \\\  
				\mathsf{if}\ \mathit{res}\ \mathsf{then}\ (l = n' \land \texttt{k} < k') \lor (r = n' \land \texttt{k} > k')\ \mathsf{else}\ (n = n' \land \texttt{k} = k')
			\end{array}
			\right\}
		}
	\end{mathpar}
	\caption{Specification of \texttt{findNext} for the binary search tree (all templates)}
	\label{fig:findNext_lock}
\end{figure*}
